Persistence in functional topology and a correction to a theorem of Morse

### 2021-07-22 surfaces.tex
Version after overall read
Version before meeting

Morse and Tompkins considered the following setting introduced by Douglas.
Let $g \colon \R \to \R^n$ be $2\pi$-periodic function representing a simple closed curve such that $g$ is differentiable with Lipschitz derivative.
Let $\widetilde{\Omega}$ be the space of continuous functions $\varphi \colon \R \to \R$ with $\varphi((t)+2\pi) = \varphi(t) + 2\pi$ for all $t$ and such that there are three distinct points $\alpha_i \in [0,2\pi)$ with $\varphi(\alpha_i)=\alpha_i$.
The \emph{Douglas functional} on $\widetilde \Omega$ associated to the curve $g$ is defined by
\begin{equation*}
A_g(\varphi)=\frac{1}{16}\int_0^{2\pi}\int_0^{2\pi}\sin\left(\frac{\alpha-\beta}{2}\right)^{-2} \! \lVert g(\varphi(\alpha))-g(\varphi(\beta)) \rVert_2^2 \ \mathrm{d}\alpha \ \mathrm{d}\beta.
\end{equation*}
Let $\Omega_g=\{\varphi\in\widetilde\Omega\mid A_g(\varphi)<\infty\}$.
Douglas proved that, for a large class of curves $g$ the set $\Omega_g$ is non-empty with compact sublevel sets of $A_g$.
Since $A_g$ is bounded below by $0$, this implies that $A_g$ has a unique global minimizer.
Douglas then used harmonic analysis to show that this minimizing critical point of $A_g$ corresponds to a solution of Plateau's Problem.

@@

Morse and Tompkins prove that each homotopically critical point of the functional $A_g$ corresponds to a critical point of the area functional, i.e., a minimal surface.
They then use a form of Morse inequalities for homotopically critical points to infer their Unstable Minimal Surface Theorem:
\begin{itemize}
    \item[($\ast$)] If $\Omega_g$ contains two distinct solutions to Plateau's Problem then it also contains an unstable minimal surface, more precisely a critical point of of $A_g$ of index 1.
\end{itemize}

@@

We have seen that these inequalities require q-tameness, so in order to apply them and deduce their theorem, Morse and Tompkins need to prove the \mbox{q-tameness} of $(\Omega_g, A_g)$.
Throughout his work on functional topology, to obtain \mbox{q-tameness} Morse assumed slightly varying forms of local-connectivity on the resulting sublevel set filtrations.
In particular, Morse and Tompkins use in their applications to minimal surface theory the following condition:
\begin{displaycquote}[p.431]{Morse.1940}
	The space $M$ is said to be \textit{locally $F$-connected} of order $r$ at $p$ if corresponding to each positive constant $e$ there exists a positive constant $\delta$ such that each singular $r$-sphere on the $\delta$-neighborhood of $p$ on $F_{c+\delta}$ bounds an $(r+1)$-cell of norm $e$ on $F_{c+e}$.
\end{displaycquote}
Here, $c = F(p)$.
Using similar language to the one used in \cref{s:connectivity}, this property is equivalent to the following notion applicable to general topological spaces.

\begin{defi}
    The sublevel set filtration of a function $F \colon M \to \R$ is said to be $\piLC$	if for any $x \in X$, $V$ a neighborhood of $x$, and any index $t > f(x)$, there is an index $s$ with  $f(x) < s < t$ and a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ induces the zero map on homotopy groups.
\end{defi}

@@

\begin{displaycquote}[Theorem 6.3, p.432]{Morse.1940}
	Let $a$ and $c$ be positive constants such that $a < c < 1$.
	The $k^{\mathrm{th}}$ connectivity $R^k(a,c)$ of $F_a$ on $F_c$ is finite.
\end{displaycquote}
Morse does not prove this statement in the given reference, but rather refers to \cite[Theorem~6.1]{Morse.1938}.
Unfortunately, this argument is not in general sound as we will show next.

\begin{defi}
    The sublevel set filtration of a function $f \colon X \to \R$ is said to be $\LC$ if for any $x \in X$, $V$ a neighborhood of $x$, and any index $t > f(x)$, there is an index $s$ with  $f(x) < s < t$ and a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is homotopic to a constant map.
\end{defi}

Clearly, being $\LC$ implies being $\piLC$ and, if the homology $\H$ takes finite dimensional values on one point spaces, also $\HLC$.
% Moreover, if the sublevel set filtration of a function $F \colon M \to \R$ on a metric space $M$ is $\LC$, then $M$ is locally $F$-connected in the sense of Morse's definition above.
We will now show that not even $\LC$ is sufficient to ensure the q-tameness of compact sublevel set filtrations.
% neither our $\HLC$ condition nor Morse's condition above are sufficient in general to imply q-tameness for compact sublevel set filtrations induced by non-continuous functions by producing an $\LC$ filtration which is not q-tame.
In particular, our construction will be used to invalidate Morse's claim quoted above.

@@

\begin{proof}
	The space $\HE$ is Hausdorff and locally compact since $\R^{d+1}$ is.
	To verify that $f$ has compact sublevel sets we notice that all level sets are either the empty set, the singleton containing the origin, or $\HE$ itself, all compact spaces.
	Let us now verify that the sublevel set filtration of $f$ is weakly $\LC$.
	Let $\epsilon > 0$ and $x \in \HE$.
	If $x$ is the origin, we choose $\delta$ between $0$ and the minimum of $\epsilon$ and $1$.
	Then $\HE_{\leq f(x) + \delta} = \HE_{\leq \delta} = \{x\}$, so the weak $\LC$ condition is trivially satisfied.
	For $x$ not the origin, there is a unique $d$-sphere in $\HE$ that contains it.
	Clearly, we may choose $\delta > 0$ so small that $D_\delta(x) = \{y \in \R^{d+1} \mid \Vert x - y \Vert < \delta\} \cap \HE$ is a disk contained in this sphere, so $D_\delta(x)$ can be contracted to $\{x\}$ and the weak $\LC$ condition again follows trivially.
	What remains to be shown is that $\HE_{\leq \bullet}$ is not q-tame for $\H$.
This follows directly from our assumption that $\H_{n}(\HE)$ is not finite dimensional for some $n$ because $\HE_{\leq t}$ is constant with value $\HE$ for $t \geq 1$.
\end{proof}

@@

\begin{cor}
    The function $f \colon \HE \to \R$ whose value at the origin is $0$ and is $1$ everywhere else defines a weakly $\LC$ compact sublevel set filtration that is not q-tame with respect to singular and \v{C}ech homology.
\end{cor}

>>>>

Morse and Tompkins considered the following setting introduced by Douglas.
Let $g \colon \R \to \R^n$ be a $2\pi$-periodic function representing a simple closed curve such that $g$ is differentiable with Lipschitz derivative.
Let $\widetilde{\Omega}$ be the space of continuous functions $\varphi \colon \R \to \R$ with $\varphi((t)+2\pi) = \varphi(t) + 2\pi$ for all $t$ and such that there are three distinct points $\alpha_i \in [0,2\pi)$ with $\varphi(\alpha_i)=\alpha_i$.
The \emph{Douglas functional} on $\widetilde \Omega$ associated to the curve $g$ is defined by
\begin{equation*}
A_g(\varphi)=\frac{1}{16}\int_0^{2\pi}\int_0^{2\pi}\sin\left(\frac{\alpha-\beta}{2}\right)^{-2} \! \lVert g(\varphi(\alpha))-g(\varphi(\beta)) \rVert_2^2 \ \mathrm{d}\alpha \ \mathrm{d}\beta.
\end{equation*}
Let $\Omega_g=\{\varphi\in\widetilde\Omega\mid A_g(\varphi)<\infty\}$.
Douglas proved that, for a large class of curves $g$ the set $\Omega_g$ is non-empty with compact sublevel sets of $A_g$.
Since $A_g$ is bounded below by $0$, this implies that $A_g$ has a unique global minimizer.
Douglas then used harmonic analysis to show that this minimizing critical point of $A_g$ corresponds to a solution of Plateau's Problem.

@@

Morse and Tompkins prove that each homotopically critical point of the functional $A_g$ corresponds to a critical point of the area functional, i.e., a minimal surface.
They then use a form of Morse inequalities for homotopically critical points to infer their Unstable Minimal Surface Theorem:
\begin{itemize}
    \item[($\ast$)] If $\Omega_g$ contains two distinct solutions to Plateau's Problem then it also contains an unstable minimal surface, more precisely a critical point of $A_g$ of index 1.
\end{itemize}

@@

We have seen that these inequalities require q-tameness, so in order to apply them and deduce their theorem, Morse and Tompkins need to prove the \mbox{q-tameness} of $(\Omega_g, A_g)$.
Throughout his work on functional topology, in order to obtain \mbox{q-tameness}, Morse assumed slightly varying forms of local connectivity on the resulting sublevel set filtrations.
In particular, in their applications to minimal surface theory, Morse and Tompkins use the following condition:
\begin{displaycquote}[p.~431]{Morse.1940}
	The space $M$ is said to be \textit{locally $F$-connected} of order $r$ at $p$ if corresponding to each positive constant $e$ there exists a positive constant $\delta$ such that each singular $r$-sphere on the $\delta$-neighborhood of $p$ on $F_{c+\delta}$ bounds an $(r+1)$-cell of norm $e$ on $F_{c+e}$.
\end{displaycquote}
Here, $c = F(p)$.
Using similar language to the one used in \cref{s:connectivity}, this property is equivalent to the following notion applicable to general topological spaces.

\begin{defi}
    The sublevel set filtration of a function $F \colon M \to \R$ is said to be \emph{homotopically locally connected} or $\piLC$	if for any $x \in X$, $V$ a neighborhood of $x$, and any index $t > f(x)$, there is an index $s$ with  $f(x) < s < t$ and a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ induces trivial maps on homotopy groups.
\end{defi}

@@

\begin{displaycquote}[Theorem 6.3, p.~432]{Morse.1940}
	Let $a$ and $c$ be positive constants such that $a < c < 1$.
	The $k^{\mathrm{th}}$ connectivity $R^k(a,c)$ of $F_a$ on $F_c$ is finite.
\end{displaycquote}
Morse does not prove this statement in the given reference, but rather refers to \cite[Theorem~6.1]{Morse.1938}.
Unfortunately, this claim does not hold in general, as we will show next by providing an explicit counterexample.

\begin{defi}
    The sublevel set filtration of a function $f \colon X \to \R$ is said to be \emph{locally connected} or $\LC$ if for any $x \in X$, $V$ a neighborhood of $x$, and any index $t > f(x)$, there is an index $s$ with  $f(x) < s < t$ and a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is homotopic to a constant map.
\end{defi}

Clearly, being $\LC$ implies being $\piLC$ and, if the homology $\H$ takes finite dimensional values on one point spaces, also $\HLC$.
We will now show that not even $\LC$ is sufficient to ensure the q-tameness of compact sublevel set filtrations that are induced by non-continuous functions.
In particular, our construction will be used to invalidate Morse's claim quoted above.

@@

\begin{proof}
    \todo{needs to be rewritten with the new s, t notation instead of the epsilon, delta notation}
	The space $\HE$ is Hausdorff %and locally compact
	since $\R^{d+1}$ is.
	To verify that $f$ has compact sublevel sets we notice that all sublevel sets are either the empty set, the singleton containing the origin, or $\HE$ itself, all compact spaces.
	Let us now verify that the sublevel set filtration of $f$ is $\LC$.
	Let $\epsilon > 0$ and $x \in \HE$.
	If $x$ is the origin, we choose $\delta$ between $0$ and the minimum of $\epsilon$ and $1$.
	Then $\HE_{\leq f(x) + \delta} = \HE_{\leq \delta} = \{x\}$, so the weak $\LC$ condition is trivially satisfied.
	For $x$ not the origin, there is a unique $d$-sphere in $\HE$ that contains it.
	Clearly, we may choose $\delta > 0$ so small that $D_\delta(x) = \{y \in \R^{d+1} \mid \Vert x - y \Vert < \delta\} \cap \HE$ is a disk contained in this sphere, so $D_\delta(x)$ can be contracted to $\{x\}$ and the weak $\LC$ condition again follows trivially.
	What remains to be shown is that $\HE_{\leq \bullet}$ is not q-tame for $\H$.
This follows directly from our assumption that $\H_{n}(\HE)$ is not finite dimensional for some $n$ because $\HE_{\leq t}$ is constant with value $\HE$ for $t \geq 1$.
\end{proof}

@@

\begin{cor}\label{c: counterexample}
    The function $f \colon \HE \to \R$ whose value at the origin is $0$ and is $1$ everywhere else defines a weakly $\LC$ compact sublevel set filtration that is not q-tame with respect to singular and \v{C}ech homology.
\end{cor}

### 2021-07-22 vietoris.tex
Version before meeting

As we have mentioned, Morse's original treatment of functional topology used metric Vietoris homology \cite{MR1512371} of filtered metric spaces. 
For greater generality we have instead considered \v{C}ech homology.
We now justify this choice showing that these two homology constructions agree on compact metric spaces.

>>>>

An interesting aspect of Morse's work from the modern point of view is his use of metric Vietoris homology with coefficients in a field, which closely relates to the use of filtered Vietoris-Rips and \v{C}ech complexes in topological data analysis.

For greater generality, instead of metric Vietoris homology we have instead considered \v{C}ech homology in the previous section.
We now justify this choice showing that these two homology constructions agree on compact metric spaces.
